From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-22 appendixB/appendixB.Rmd
WIP

---
title: "Appendix B: System size expansion of the master equation for demographic noise"
author: 
  name: "Carl Boettiger"
  affiliation: a
address:
  - code: a
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
layout: 5p  
bibliography: ../paper/refs.bib
output: rticles::elsevier_article
preamble: "\\newcommand{\\ud}{\\mathrm{d}}"
---

@@

The first integral (over $\textrm{d}t$) is the familiar Reimann integral of the classic calculus, but the second is an Itô integral, which integrates over the stochastic process $B_t$ (Brownian motion, also termed a Weiner process). The Itô calculus follows different rules from the classic calculus, largely due to the fact that an Itô integral has non-zero quadratic variation. (Crudely put, $\textrm{d}x^2 \sim 0$, but $\textrm{d}B^2 \sim dx$).

The intepretation of Langevin equation is less precise, as it can be read to correspond to one of two possible stochastic integrals, and is often associated with a Stratonovich integral.  (Unlike the Itô integral, the Stratonovich integral has zero quadratic variation and so obeys the same chain rule as the classic calculus, but it is not a martingale, e.g. an unbiased random walk, which is a convenient for property for proving theorems).  Fortunately, the notion as a partial differential equation over probabilities does not have this ambiguity, which shows that @Ovaskainen2010 is referring to an Itô expression, just as I have written it more explicitly.  See @vanKampen2007 for an excellent discussion of the Itô-Stratonovich dillema and its intepretation with respect to intrinsic vs environmental noise factors.

## Master equations and the step operator

Consider a population that is of size $n$ with probability $p_n$, with birth and death rates given by $b_n$ and $d_n$ respectively, diagrammed in Fig~\ref{markov}  This representation is known as a single step Markov process.  The rate of transitions to the $n+1$ state $b_n$, the rate of transitions to $n-1$ is $d_n$.  Both of these are transitions away from the state $p_n$, hence the decrease the probability of $p_n$.  The probability $p_n$ increases by transitions into the state $n$, from either side: births enter from the state below at rate $b_{n-1}$ and deaths from the state above, $d_{n+1}$.  Hence the rate of change in $p_n$ is given by

@@

## A traditional derivation

A derivation of the above formulation can be found, e.g. in @Nisbet1982's classic textbook on fluctuating populations, or more recently in @Lande2003.  This diffusion approximation is also given for a more mathematically inclined audience in @Gardiner2009 and in @vanKampen2007.  The approach starts with the same master equation that defines the underlying dynamics of births and deaths (matching the process implemented in the exact Gillespie algorithm, @Gillespie1977, see Appendix A).

@@

## WKB

@Ovaskainen2010 provides an overview literature in that addresses this limit and discusses the use of the WKB^[Wentzel, Kramers & Brillouin. Naturally this is KWB in Holland, BWK in France, and JWKB in England (for Jeffreys). Note that the physicist Gregor Wentzel is not the same as the mathematician Alexander Wentzell of the Fredilin-Wentzell theorem for large deviations years later, though I believe the former's result can be derived directly from the latter's.] method to address this case.

>>>>

---
title: "Appendix B: System size expansion of the master equation for demographic noise"
author: 
  name: "Carl Boettiger"
  affiliation: a
address:
  - code: a
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
layout: 5p  
bibliography: ../paper/refs.bib
output: 
  rticles::elsevier_article:
    toc: true

preamble: "\\newcommand{\\ud}{\\mathrm{d}}"
---

\renewcommand{\thefigure}{B\arabic{figure}}

# Preliminary comments on notation

@@

The first integral (over $\textrm{d}t$) is the familiar Reimann integral of the classic calculus, but the second is an Itô integral, which integrates over the stochastic process $B_t$ (Brownian motion, also termed a Weiner process). The Itô calculus follows different rules from the classic calculus, largely due to the fact that an Itô integral has non-zero quadratic variation. (Crudely put, $\textrm{d}x^2 \sim 0$, but $\textrm{d}B^2 \sim dx$).  The SDE notation also traditionally adopts capital letters such as $X_t$ to denote stochastic variables.

The intepretation of Langevin equation is less precise, as it can be read to correspond to one of two possible stochastic integrals, and is often associated with a Stratonovich integral.  (Unlike the Itô integral, the Stratonovich integral has zero quadratic variation and so obeys the same chain rule as the classic calculus, but it is not a martingale, e.g. an unbiased random walk, which is a convenient for property for proving theorems).  Fortunately, the notion as a partial differential equation over probabilities does not have this ambiguity, which shows that @Ovaskainen2010 is referring to an Itô expression, just as I have written it more explicitly.  See @vanKampen2007 for an excellent discussion of the Itô-Stratonovich dillema and its intepretation with respect to intrinsic vs environmental noise factors. To avoid this ambiguity, mathematical literature will frequently write the Statonovich-type SDE explicitly as $\mathrm{d} X_t = f(X_t) \mathrm{d}t +  g(X_t) \circ  B_t$

### Partial Differential Equation formulation

An SDE can also be written in terms of an equivalent partial differential equation for the probability distribution of a given variable.  The SDE

\begin{equation}
\textrm{d} X_t = \mu(X_t) \textrm{d}t + \sigma(X_t) \textrm{d}B_t
\end{equation}

is equivalent to the PDE formulation [e.g. @vanKampen2007 or @Oksendal1985].

\begin{equation}
\frac{\partial}{\partial t} P(x,t) = - \frac{\partial}{\partial x} \mu(x) P(x,t) + \frac{1}{2} \frac{\partial^2}{\partial x^2} \sigma(x)^2 
\end{equation}

# Master equations and the step operator

Consider a population that is of size $n$ with probability $p_n$, with birth and death rates given by $b_n$ and $d_n$ respectively, diagrammed in Fig \ref{markov}  This representation is known as a single step Markov process.  The rate of transitions to the $n+1$ state $b_n$, the rate of transitions to $n-1$ is $d_n$.  Both of these are transitions away from the state $p_n$, hence the decrease the probability of $p_n$.  The probability $p_n$ increases by transitions into the state $n$, from either side: births enter from the state below at rate $b_{n-1}$ and deaths from the state above, $d_{n+1}$.  Hence the rate of change in $p_n$ is given by

@@

## The system-size expansion vs the diffusion approximation

@Kramers1940 @Moyal1949, @Kurtz1978

@vanKampen1961 @vanKampen1976 @Kurtz1972 @Kurtz1978

A derivation of diffusion approximation in @Nisbet1982's classic textbook on fluctuating populations, or more recently in @Lande2003.  This diffusion approximation is also given for a more mathematically inclined audience in @Gardiner2009 and in @vanKampen2007.  The approach starts with the same master equation that defines the underlying dynamics of births and deaths (matching the process implemented in the exact Gillespie algorithm, @Gillespie1977, see Appendix A).

@@

## Large deviations

Notably, both the SDE derived from the Kramer-Moyal diffusion approximation and the central limit theorem result arising from the system size expansion break down in the case of large deviations.  @Ovaskainen2010 provides an overview literature in that addresses this limit and discusses the use of the WKB^[Wentzel, Kramers & Brillouin. Note that the physicist Gregor Wentzel is not the same as the mathematician Alexander Wentzell of the Fredilin-Wentzell theorem for large deviations years later, though I believe the former's result can be established more formally by the theorem from the latter.] method to address this case.

### 2018-02-22 paper/paper.Rmd
WIP

Where $\textrm{d}B_t^{(d)}$ and $\textrm{d}B_t^{(e)}$ refer to Brownian processes, (i.e. Gaussian white noise process) for demographic and environmental stochasticity respectively.  When $f(n)$ is simply a Verhulst logistic function this is often referred to as the canonical model [e.g. @@Ovaskainen2010], and provides a clear illustration of the partition of separate deterministic and stochastic elements. Significant research continues to focus on extending the study of this model to consider more complex and higher-dimensional $f(n)$ (coexistence models, stage structured models) and consider the case of autocorrelated, rather than than white noise of $\textrm{d}B_t^{(e)}$.

Before we follow that path, I will review how a shift towards more mechanistic descriptions of the origins of noise provide a more nuianced view of when the canonical model does an does not hold in the context of both demographic and environmental stochasticity.  My approach will highlight two fundamental tools: an ability for efficient, exact simulation using @Gillespie1977's stochastic simulation algorithm (SSA), and the van Kampen system size expansion [@vanKampen2007], which is rooted in the mathematical theorems of @Kurtz1970 & @Kurtz1971, in place of the more popular diffusion approximation derived by Taylor series [e.g. @Nisbet1982 @Lande2003; see comparisons in @vanKampen2007 and @Gardiner2009].

@@

The van Kampen system size expansion, shown for a general birth-death
process in Appendix B, derives an ordinary differential equation (ODE)
for the macroscopic process and a *linear* Fokker-Planck equation which
governs the *deviations* from the macroscopic equation:

\begin{align}
\frac{\mathrm{d} x }{dt}  &= b(x) - d(x) + O(N^{-1})\\
\frac{\mathrm{d} \sigma^2 }{dt}  &= 2 \left( \partial_x b(x)  - \partial_x d(x) \right) \sigma^2 + b(x) + d(x) \label{bd}
\end{align}

where $x = \langle n \rangle$, the average or expected population
size of an ensemble of replicates, assuming the transitions (+/- 1
in this case) are much smaller than the system size ($N$ in the
Levins model).  The dynamics for the average population
match the population-level equation.^[up to fluctuations of order
$\tfrac{1}{N}$. We can actually both observe and successfully
approximate this higher-order correction from additional terms in
the system size expansion to recover the phenomenon of stochastic
inflation [@Gonzalez2002], where the average population size can
be larger or smaller than predicted by the deterministic model --
see Appendix A for examples in simulations compared to predictions
of the system size expansion].

Note that the van Kampen expansion is not identical to the SDE 
postulated by the canonical equation, which corresponds to a 
non-linear Fokker-Planck equation for the probability density.  
As shown in Appendix B, the popular expansion

@@

\begin{equation}
\sigma_x^2  = \frac{ \left( \partial_y f(x,y) \right)^2}{\left( \partial_x f(x,y) \right)^2 + \partial_x f(x,y) \tau_c } \sigma_e^2 + \sigma^2_d \label{env_noise}
\end{equation}

>>>>

Where $\textrm{d}B_t^{(d)}$ and $\textrm{d}B_t^{(e)}$ refer to Brownian processes, (i.e. Gaussian white noise process) for demographic and environmental stochasticity respectively.  When $f(n)$ is simply a Verhulst logistic function this is often referred to as the canonical model [e.g. @Ovaskainen2010], and provides a clear illustration of the partition of separate deterministic and stochastic elements. Significant research continues to focus on extending the study of this model to consider more complex and higher-dimensional $f(n)$ (coexistence models, stage structured models) and consider the case of autocorrelated, rather than than white noise of $\textrm{d}B_t^{(e)}$.

Before we follow that path, I will highlight

I will review how a shift towards more mechanistic descriptions of the origins of noise provide a more nuianced view of when the canonical model does an does not hold in the context of both demographic and environmental stochasticity.  My approach will highlight two fundamental tools: an ability for efficient, exact simulation using @Gillespie1977's stochastic simulation algorithm (SSA), and the van Kampen system size expansion [@vanKampen1976].

which is rooted in the mathematical theorems of Thomas Kurtz [@Kurtz1970 & @Kurtz1971, @Kurtz1972 @Kurtz1978], in place of the more popular diffusion approximation derived by Taylor series [e.g. @Nisbet1982, @Lande2003; rooted in the earlier work of @Kramers1940 and @Moyal1949, see comparisons in @vanKampen2007 and @Gardiner2009].

@@

The van Kampen system size expansion, shown for a general birth-death
process in Appendix B, derives an *ordinary* differential equation (ODE)
for the macroscopic process, coupled to a *linear* SDE which
governs the *deviations* ($\xi$)  from the macroscopic equation:

\begin{align}
\frac{\mathrm{d} x }{dt}  &= b(x) - d(x) + \mathcal{O}(N^{-1}) \\
\mathrm{d} \xi &=  \partial_{x}  \left( b(x)  - d(x) \right)  \xi \mathrm{d}t + \sqrt{b(x) + d(x)} \mathrm{d} B_t   + \mathcal{O}(N^{-\tfrac{1}{2}}
\end{align}

\begin{align}
\frac{\mathrm{d} x }{\mathrm{d} t}  &= b(x) - d(x) + \mathcal{O}(N^{-1})\\
\frac{\mathrm{d} \sigma^2 }{\mathrm{d} t}  &= 2 \left( \partial_x b(x)  - \partial_x d(x) \right) \sigma^2 + b(x) + d(x)  + \mathcal{O}(N^{-\tfrac{1}{2}}) \label{bd}
\end{align}

The dynamics for the average population
match the population-level equation.^[up to fluctuations of order
$\tfrac{1}{N}$. We can actually both observe and successfully
approximate this higher-order correction from additional terms in
the system size expansion to recover the phenomenon of stochastic
inflation [@Gonzalez2002], where the average population size can
be larger or smaller than predicted by the deterministic model --
see Appendix A for examples in simulations compared to predictions
of the system size expansion].

Note that the van Kampen expansion is not identical to the SDE 
postulated by the-so called canonical equation.  
As shown in Appendix B, the popular expansion

@@

$$
\begin{equation}
\sigma_x^2  = \frac{ \left( \partial_y f(x,y) \right)^2}{\left( \partial_x f(x,y) \right)^2 + \partial_x f(x,y) \tau_c } \sigma_e^2 + \sigma^2_d 
\label{env_noise}
\end{equation}
$$
